---
title: "Cleaning RMD"
author: "Lim Ryan"
date: "2023-05-04"
output: html_document
---

# Cleaning Process in R

Each month's data table is cleaned by doing the following:

- Remove Duplicated Ride Id
- Remove Inconsistent Started At and Ended At Timings
- Remove Incomplete Data Rows
- Validate rideable_type, member_casual Columns
- Adding day_type Column

The cleaned files will be saved as Format of cleaned data: clean_[yyyy]_[mm]

## Packages to Install
```{r}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
install.packages("readxl")
install.packages("dplyr")
install.packages("tidyverse")
install.packages("geosphere")

library(readxl)
library(dplyr)
library(tidyverse)
library(geosphere)
```

## Combining 12 Months of Historic Data
> Creating a variable to store the list of files and looping through it to combine the 12 months of historic data
EDIT PATH LATER

```{r}
csv_dir <- "D:/Job/Google Case Study/"

csv_files <- list.files(csv_dir, pattern = "\\.csv$")
all_data <- data.frame()
for (file in csv_files) {
  file_path <- file.path(csv_dir, file)
  data <- read_csv(file_path, col_types = cols())
  all_data <- bind_rows(all_data, data)
  }
```

## Adding Data Frame into Environment
> Adding all_data frame into a clean_12_months where all the cleaning and manipulation is done

```{r} 
clean_12_months <- all_data
```

## Remove Duplicated Ride Id
> Group by ride id and slicing it by 1 will remove any duplicated ride id

```{r}
clean_12_months <- clean_12_months %>% 
     group_by(ride_id) %>% 
     slice(1)
```
	 
## Removing inconsistent Dates
> Inconsistent timings: values that are negative or below 1 after "ended_at" - "started_at"
> Using vectorisation to speed up the processing of huge data
> Converting to datetime format & adding in a trip_length column

Vectorisation is applying a function or operation to an entire vector or column of data at once, rather than looping through the data row by row.

```{r}
clean_12_months$started_at <- as.POSIXct(clean_12_months$started_at, format = "%Y-%m-%d %H:%M:%S")
clean_12_months$ended_at <- as.POSIXct(clean_12_months$ended_at, format = "%Y-%m-%d %H:%M:%S")
clean_12_months$trip_length <- as.numeric(difftime(clean_12_months$ended_at, clean_12_months$started_at, units = "mins"))
clean_12_months <-subset(clean_12_months, trip_length >= 1)
```

## Removing Incomplete Data Rows
> Any fields with NA will be completely omitted
```{r}
clean_12_months <- na.omit(clean_12_months)
```

## Validate rideable_type, member_casual Columns
> Contain 3 types of bikes: electric, docked, classic
> Contain 2 types of people: member, casual

```{r}
clean_12_months <- clean_12_months %>% 
  filter(rideable_type %in% c("electric_bike", "docked_bike", "classic_bike")) %>%
  filter(member_casual %in% c("member", "casual"))
```

## Validate start_lat, start_lng, end_lat, end_lng Columns
> Removing those that are 0.0000. It is an impossible value, in the middle of the ocean

```{r}
clean_12_months <- clean_12_months %>% 
  filter(start_lat != 0 & start_lng != 0 & end_lat != 0 & end_lng != 0)
```

## Adding day_type Column
> Identifying the day of the ride using weekdays function from R

```{r}
clean_12_months$day_of_week <- as.integer(factor(weekdays(clean_12_months$started_at),
                                                 levels = c("Sunday", "Monday", "Tuesday",
                                                            "Wednesday", "Thursday", "Friday",
                                                            "Saturday")))
```

## Adding Distance Column
> Calculating distance using start and end points

```{r}
clean_12_months$distance <- distHaversine(clean_12_months[c("start_lng", "start_lat")], 
                                      clean_12_months[c("end_lng", "end_lat")])
```

## Overview of Cleaned Data
```{r}
summary(clean_12_months)
```


